Linting Linearity in Core/System FC

### 2022-11-18 proposal.tex
Fix another example using the case binder unifying solution
Rewrite CSE and Desugarer examples

\section{Motivation}

@@

\subsection{Result of desugar (before optimization)}

This resulting program fails lint with \emph{Linearity failure in lambda: x
'Many $\not\subseteq$ p}
\begin{code}
($!) :: forall {rep} a (b :: TYPE rep) p q. (a %p -> b) %q -> a %p -> b
($!) f !x = f x
~~>
($!)
  :: forall {rep :: RuntimeRep} a (b :: TYPE rep) (p :: Multiplicity)
            (q :: Multiplicity).
     (a %p -> b) %q -> a %p -> b
($!)
  = \ (@(rep :: RuntimeRep))
      (@a)
      (@(b :: TYPE rep))
      (@(p :: Multiplicity))
      (@(q :: Multiplicity))
      (f :: a %p -> b)
      (x :: a) ->
      case x of x { __DEFAULT -> f x }
\end{code}

How are strictness annotations typed in the frontend? The issue is this program
consumes to value once to force it, and then again to determine the return
value.

Work better the meaning of consumption (does it mean for a value to be reduced
to WHNF?). Is consuming = forcing? Why is the above program linear?

The resulting program from -ddump-simpl is
\begin{code}
($!)
  :: forall {rep :: GHC.Types.RuntimeRep} a (b :: TYPE rep)
            (p :: GHC.Types.Multiplicity) (q :: GHC.Types.Multiplicity).
     (a %p -> b) %q -> a %p -> b
($!)
  = \ (@(rep :: GHC.Types.RuntimeRep))
      (@a)
      (@(b :: TYPE rep_aSJ))
      (@(p :: GHC.Types.Multiplicity))
      (@(q :: GHC.Types.Multiplicity))
      (f :: a %p -> b)
      (x :: a) ->
      case x of y { __DEFAULT -> f_aDV y }
\end{code}
So in this program we use another name for the case binder, but it still stands
for the resulting of evaluating to WHNF; how is it technically different?

I don't understand this one yet. I would guess it has to do with multiplicity
coercions. Incorrect incompatible branch: CoAxBranch (src/Prelude/Linear/GenericUtil.hs:112:3-51): [S1
\begin{code}
type family Fixup (f :: Type -> Type) (g :: Type -> Type) :: Type -> Type where
  Fixup (D1 c f) (D1 _c g) = D1 c (Fixup f g)
  Fixup (C1 c f) (C1 _c g) = C1 c (Fixup f g)
  Fixup (S1 c f) (S1 _c (MP1 m f)) = S1 c (MP1 m f) -- error in this constructor
  Fixup (S1 c f) (S1 _c f) = S1 c f
  Fixup (f :*: g) (f' :*: g') = Fixup f f' :*: Fixup g g'
  Fixup (f :+: g) (f' :+: g') = Fixup f f' :+: Fixup g g'
  Fixup V1 V1 = V1
  Fixup _ _ = TypeError ('Text "FixupMetaData: representations do not match.")
\end{code}

Example
\begin{code}
(&&) :: Bool %1 -> Bool %1 -> Bool
False && False = False
False && True = False
True && x = x
~~>
(&&) :: Bool %1 -> Bool %1 -> Bool
(&&)
  = \ (x :: Bool) (y :: Bool) ->
      case x of w1 {
        False -> case y of w2 { __DEFAULT -> x };
        True -> y
      }
\end{code}
In the first branch case, we pattern match on y to force it and then return $x$
rather than the constant $False$.

At first look, the resulting program is impossible to typecheck.

Idea: if $x$ was anotated with some information regarding the CSE then perhaps
it could be typechecked (in a system that considered said anotations) -- This is
too specific and the case binder environment solution is much cleaner. It just
requires that the example is changed into using the case binder $w1$ rather than
$x$.

Solution with the (unifying) case binder idea:
Annotate the case binder with its usage environment. For enums without arguments
this usage environment will always be empty, meaning we can always use the case
binder instead of the actual enum without issues. This solves this problem as
well as a handful of other.

In this exact example, the binder has usage environment empty ($w1 := []$),
meaning that in the False branch, when we use $x$, if we instead used $w1$ which
is equivalent and seems more correct (since it doesn't need the case scrutinee
expression to be a variable), then the usage of $w1$ would imply the usage of
$[]$ which is nothing and therefore we would preserve linearity

Confusingly: the resulting optimised program from running the complete simplifier
and outputting with -ddump-simpl actually does what we expected it to regarding
the constants and linearity. So is the issue from running linear lint after a
particular CSE but it would be fine in the end?
\begin{code}
(&&) :: Bool %1 -> Bool %1 -> Bool
(&&)
  = \ (x :: Bool) (y :: Bool) ->
      case x of {
        False -> case y of { __DEFAULT -> GHC.Types.False };
        True -> y
      }
\end{code}

@@

@@

>>>>

\subsection{Motivation}

@@

@@

\section{Novel rules}

\subsection{The usage environment in a Recursive Let}

\subsection{The usage environment of a Case Binder\label{caseBinderKey}}

\begin{mathparpagebreakable}
    \infer*[right=(letrec)]
    {\Gamma ; \Delta/ \Delta' ; \Omega \vdash M : A \Uparrow \and \Gamma ;
    \Delta/ \Delta'' ; \Omega \vdash N : B \Uparrow \and \Delta' = \Delta''}
    {\Gamma ; \Delta/\Delta' ; \Omega \vdash  (M \with N) : A \leadsto U}
\end{mathparpagebreakable}

\section{Examples}

In this section we present worked examples of programs that currently fail to
compile with \textbf{-dlinear-core-lint} but would succeed according to the
novel typing rules we introduced above. Each example belongs to a subsection
that indicates after which transformation the linting failed.

\subsection{After the desugarer (before optimizations)}

The definition for $\$!$ in \textbf{linear-base}\cite{} fails to lint after the
desugarer (before any optimisation takes place) with \emph{Linearity failure in
lambda: x 'Many $\not\subseteq$ p}
\begin{code}
($!) :: forall {rep} a (b :: TYPE rep) p q. (a %p -> b) %q -> a %p -> b
($!) f !x = f x
\end{code}

The desugared version of that function follows below. It violates (naive?)
linearity by using $x$ twice, once to force the value and a second time to call
$f$. However, the $x$ passed as an argument is actually the case binder and it
must be handled in its own way. The case binder is the key (as seen in ~\ref{caseBinderKey}) to solving this
and many other examples.
\begin{code}
($!)
  :: forall {rep :: RuntimeRep} a (b :: TYPE rep) (p :: Multiplicity)
            (q :: Multiplicity).
     (a %p -> b) %q -> a %p -> b
($!)
  = \ (@(rep :: RuntimeRep))
      (@a)
      (@(b :: TYPE rep))
      (@(p :: Multiplicity))
      (@(q :: Multiplicity))
      (f :: a %p -> b)
      (x :: a) ->
      case x of x { __DEFAULT -> f x }
\end{code}

% Intuitively, this program is linear because forcing the polymorphic value to
% WHNF ...? Need better semantics of consuming
%
The case binder usage rule typechecks this program because $x$ is consumed once,
the usage environment of the case binder is empty ($x :_\emptyset a$) and,
therefore, when $x$ is used in $f(x)$, we use its usage environment which is
empty, so we don't use anything new.

\emph{How are strictness annotations typed in the frontend? The issue is this program
consumes to value once to force it, and then again to determine the return
value.}

% Work better the meaning of consumption (does it mean for a value to be reduced
% to WHNF?). Is consuming = forcing? Why is the above program linear?

As a finishing note on this example, we show the resulting program from
\textbf{-ddump-simpl} that uses a different name for the case binder.
\begin{code}
($!)
  :: forall {rep :: RuntimeRep} a (b :: TYPE rep)
            (p :: Multiplicity) (q :: Multiplicity).
     (a %p -> b) %q -> a %p -> b
($!)
  = \ (@(rep :: RuntimeRep))
      (@a)
      (@(b :: TYPE rep_aSJ))
      (@(p :: Multiplicity))
      (@(q :: Multiplicity))
      (f :: a %p -> b)
      (x :: a) ->
      case x of y { __DEFAULT -> f_aDV y }
\end{code}
% So in this program we use another name for the case binder, but it still stands
% for the resulting of evaluating to WHNF; how is it technically different?

The definition for $\&\&$ in \textbf{linear-base} fails to lint after the common
sub-expression elimination (CSE) pass transforms the program.
\begin{code}
(&&) :: Bool %1 -> Bool %1 -> Bool
False && False = False
False && True = False
True && x = x
\end{code}
The issue with the program resulting from the transformation is $x$ being used
twice in the first branch of the first case expression. We pattern match on y to
force it (since it's a type without constructor arguments, forcing is consuming)
and then return $x$ rather than the constant $False$
\begin{code}
(&&) :: Bool %1 -> Bool %1 -> Bool
(&&) = \ (x :: Bool) (y :: Bool) ->
  case x of w1 {
    False -> case y of w2 { __DEFAULT -> x };
    True -> y
  }
\end{code}
At a first glance, the resulting program is impossible to typecheck linearly.

The key observation is that after $x$ is forced into either $True$ or $False$,
we know $x$ to be a constructor without arguments (which can be created
without restrictions) and know that where we see $x$ we can as well have
$True$ or $False$ depending on the branch. However, using $x$ here is very
unsatisfactory (and linearly unsound?) because $x$ might be an expression, and
we can't really associate $x$ to the value we pattern matched on (be it $True$
or $False$). What we really want to have instead of $x$ is the $w1$ case binder --
it relates directly to the value we pattern matched on, it's a variable rather
than an expression, and, most importantly, our case-binder-usage idea could be
applied here as well

% Idea: if $x$ was anotated with some information regarding the CSE then perhaps
% it could be typechecked (in a system that considered said anotations) -- This is
% too specific and the case binder environment solution is much cleaner. It just
% requires that the example is changed into using the case binder $w1$ rather than
% $x$.

The solution with the unifying case binder usage idea means annotating the case
binder with its usage environment. For $True$ and $False$ (and other data
constructors without arguments) the usage environment is always empty (using
them doesn't entail using any other variable), meaning we can always use the
case binder instead of the actual constant constructor without issues.
%
In concrete, if we had $w1$ instead of the second occurrence of $x$, we'd have
an empty usage environment for $w1$ in the $False$ branch ($w1 :_\emptyset
Bool$) and upon using $w1$ we wouldn't use any extra resources

To make this work, we'd have to look at the current transformations and see
whether it would be possible to ensure that CSEing the case scrutinee would
entail using the case binder instead of the scrutinee. I don't know of a
situation in which we'd really want the scrutinee rather than the case binder,
so I hypothetize the modified transformation would work out in practice and
solve this linearity issue.

% In this exact example, the binder has usage environment empty ($w1 := []$),
% meaning that in the False branch, when we use $x$, if we instead used $w1$ which
% is equivalent and seems more correct (since it doesn't need the case scrutinee
% expression to be a variable), then the usage of $w1$ would imply the usage of
% $[]$ which is nothing and therefore we would preserve linearity

Curiously, the optimised program resulting from running all transformations
actually does what we expected it with regard to using the constant constructors
and preserving linearity. So is the issue from running linear lint after a
particular CSE but it would be fine in the end?
\begin{code}
(&&) :: Bool %1 -> Bool %1 -> Bool
(&&) = \ (x :: Bool) (y :: Bool) ->
  case x of {
    False -> case y of { __DEFAULT -> GHC.Types.False };
    True -> y
  }
\end{code}

@@

I don't understand this one yet. I would guess it has to do with multiplicity
coercions. Incorrect incompatible branch: CoAxBranch (src/Prelude/Linear/GenericUtil.hs:112:3-51):
\begin{code}
type family Fixup (f :: Type -> Type) (g :: Type -> Type) :: Type -> Type where
  Fixup (D1 c f) (D1 _c g) = D1 c (Fixup f g)
  Fixup (C1 c f) (C1 _c g) = C1 c (Fixup f g)
  Fixup (S1 c f) (S1 _c (MP1 m f)) = S1 c (MP1 m f) -- error in this constructor
  Fixup (S1 c f) (S1 _c f) = S1 c f
  Fixup (f :*: g) (f' :*: g') = Fixup f f' :*: Fixup g g'
  Fixup (f :+: g) (f' :+: g') = Fixup f f' :+: Fixup g g'
  Fixup V1 V1 = V1
  Fixup _ _ = TypeError ('Text "FixupMetaData: representations do not match.")
\end{code}
